\chapter{Model and Training}
\label{ch:model}

\section{Overview}

We extend the existing \texttt{TurnTransformer} backbone with auxiliary belief
heads and optional belief fusion into the policy head. The design keeps
parameter count modest (on the order of one to five million) for Master's-scale
compute.

\section{Hierarchical Encoder}

\subsection{Within-Turn Self-Attention}

For each turn $t$, token embeddings are summed with slot position embeddings and
passed through a transformer encoder over the thirteen slots:
\[
  h_t = \mathrm{MeanPool}\bigl(\mathrm{SlotEnc}(E(x_t) + P_\mathrm{slot})\bigr).
\]
Mean pooling over slot outputs yields one turn embedding $h_t \in \mathbb{R}^d$.

\subsection{Cross-Turn Causal Attention}

Turn embeddings receive turn position embeddings and pass through a causal
transformer encoder:
\[
  \tilde{h}_t = \mathrm{TurnEnc}_\mathrm{causal}(h_{1:t} + P_\mathrm{turn})_t.
\]
The representation at the last valid turn index is used for prediction heads.

This two-level structure lets slot attention relate field state to individual
party members within a turn while turn attention integrates evidence across
time without peeking at future observations.

\section{Policy Head}

The policy head is a linear classifier over the final turn embedding:
\[
  \pi(a \mid C_t) = \mathrm{softmax}(W_\pi \tilde{h}_{t^*}),
\]
where $t^*$ is the index of the most recent valid turn. Training minimizes
cross-entropy against the human action (behavioral cloning).

\section{Belief Heads}

For each belief target type $k$ and opponent slot $s$, a small MLP maps
$\tilde{h}_{t^*}$ to logits over the target vocabulary:
\[
  \hat{z}_{t,s}^{(k)} = \mathrm{softmax}\bigl(f_s^{(k)}(\tilde{h}_{t^*})\bigr).
\]
Species prediction uses six slot-specific heads sharing the same trunk
representation. Masked slots exclude terms from the belief loss.

\section{Belief Fusion Variant}

In an ablation we concatenate belief head softmax outputs to the policy input:
\[
  \pi(a \mid C_t) = \mathrm{softmax}\bigl(W_\pi [\tilde{h}_{t^*}; \hat{z}_t]\bigr).
\]
This tests whether explicit belief features help action selection beyond shared
representation learning in the trunk.

\section{Training Procedure}

\subsection{Stage A: Belief-Augmented Behavioral Cloning}

The multi-task loss is
\begin{equation}
  \mathcal{L} = \mathcal{L}_\mathrm{action}
    + \lambda_\mathrm{bel} \sum_k \mathcal{L}_\mathrm{bel}^{(k)},
  \label{eq:multitask}
\end{equation}
where $\mathcal{L}_\mathrm{action}$ is action cross-entropy and each
$\mathcal{L}_\mathrm{bel}^{(k)}$ is masked cross-entropy over opponent slots.
The weight $\lambda_\mathrm{bel}$ is tuned on validation data.

\subsection{Stage B: Conservative Offline RL Fine-Tuning}

We initialize from Stage A and optionally fine-tune with CQL or AMAGO on the
same offline trajectories. No search or simulator rollouts run at inference.

\subsection{Stage C: Evaluation}

All benchmark battles use frozen checkpoints. No online weight updates occur
during evaluation.

\section{Baselines}

Belief baselines include gradient-boosted trees and logistic regression on
hand-crafted features, MLP on the latest turn only, GRU over turn embeddings,
TurnTransformer without belief loss, and a flat causal transformer over
$13 \times T$ tokens.

Policy baselines include BC with MLP, BC with GRU, BC TurnTransformer, belief-
augmented TurnTransformer, and belief-augmented models with offline RL
fine-tuning.

\section{Architecture Summary}

\begin{table}[ht]
  \centering
  \caption{Default hyperparameters for the belief transformer.}
  \label{tab:hyperparams}
  \begin{tabular}{ll}
    \toprule
    Hyperparameter & Default \\
    \midrule
    Model dimension $d$ & 128 \\
    Attention heads & 4 \\
    Slot encoder layers & 1 \\
    Turn encoder layers & 2 \\
    Context turns $N$ & 8 \\
    Dropout & 0.1 \\
    Optimizer & AdamW, lr $3\times10^{-4}$ \\
    $\lambda_\mathrm{bel}$ & tuned on validation \\
    \bottomrule
  \end{tabular}
\end{table}

The \texttt{BeliefTransformer} class implements belief heads and optional fusion.
Training is invoked via \texttt{scripts/train\_belief\_bc.py}.
