\chapter{Experimental Setup}
\label{ch:experiments}

\section{Dataset}

The primary corpus is Metamon Gen~9 OU human replays. Phase~1 trains turn-only
behavioral cloning on an encoded subset of approximately 255\,000 turns.
Phase~2 exports a belief-enriched corpus of 2\,000 battles (69\,411 turns) with
active-opponent move, item, ability, and tera labels plus team matchup keys.
Observations use first-person reconstruction from spectator logs. Actions come
from human moves. Belief labels come from full-state opponent annotations that
are unavailable to the agent at decision time.

Random-bot logs from poke-env support pipeline validation before human data is
used for research runs.

\section{Data Splits}

We report two primary split strategies and defer a third.

\subsection{Split A: Random Battle Split}

Standard 80/20 train and validation partition by battle ID. This measures
in-distribution imitation and belief accuracy on unseen battle trajectories.

\subsection{Split B: Held-Out Team Cores}

Each opponent team receives a canonical key from sorted species names. Teams in
a designated held-out set never appear in training. Evaluation compares turns
whose opponent team key appeared in training (ID train teams) against turns
whose key was unseen (OOD teams). The generalization gap is ID train team
accuracy minus OOD team accuracy and is the primary summary statistic for RQ3.

\subsection{Split C: Held-Out Set Combinations (Deferred)}

Holding out specific tuples of item, ability, and move quartet would test
finer combinatorial shift beyond species identity. Split~C is left for future
work because Gen~9 OU already reveals species at team preview and because the
scaled Phase~2 results already answer the species-key version of RQ3.

\section{Belief Metrics}

For move-slot belief we report top-1 accuracy on masked unrevealed slots.
Baselines include a latest-turn MLP, a GRU over eight turns, and a linear probe
on a frozen BC trunk. Item, ability, and tera heads are trained jointly with
move heads. Aggregate move-slot accuracy is the primary RQ1 number reported in
\Cref{ch:results}. Calibration plots (ECE, Brier) and reveal-turn anticipation
curves are deferred.

\section{Policy Metrics}

Policy evaluation uses
\begin{itemize}
  \item BC cross-entropy and top-1 action match versus human actions on Split~A,
  \item Split~B ID versus OOD team action match and the generalization gap,
  \item ablation comparisons among turn-only BC, matchup BC, ActiveBelief, and
    belief-fusion variants.
\end{itemize}

Live win rate on poke-env is an optional play-strength metric. It is not required
to answer RQ1--RQ3 as framed in this thesis. Offline action match measures
imitation fidelity, not ladder competitiveness against humans.

\section{Ablations}

\begin{table}[ht]
  \centering
  \caption{Ablation studies executed in Phase~2.}
  \label{tab:ablations}
  \begin{tabular}{ll}
    \toprule
    Ablation & What it tests \\
    \midrule
    Turn-only BC vs Matchup BC & Value of explicit team encoder \\
    Matchup BC vs ActiveBelief & Value of auxiliary belief heads \\
    Belief fusion vs parallel heads & Explicit belief features for policy \\
    MLP / GRU / linear probe vs ActiveBelief & RQ1 baseline comparison \\
    500 vs 2\,000 battle scale & Data coverage versus architecture \\
    \bottomrule
  \end{tabular}
\end{table}

Context-window sweeps, flat $13T$ sequence transformers, and offline RL
(CQL or AMAGO) remain future work.

\section{Statistical Reporting}

We report point estimates on fixed validation and OOD splits shared across
models. Paired comparisons use the same encoded corpus and token vocabulary.
Bootstrap confidence intervals over battles are desirable for a journal revision
but are not required for the Master's-scale tables in \Cref{ch:results}.

\section{Minimum Viable Result}

The thesis meets its pass bar if at least two of the following hold.
\begin{enumerate}
  \item Belief models demonstrate nontrivial move-belief accuracy relative to
    chance and beat at least one strong neural baseline on that target.
  \item Belief-augmented BC improves OOD team metrics or shrinks the
    generalization gap relative to turn-only BC.
  \item Offline RL fine-tuning improves offline or live policy metrics over
    BC-only without search at inference.
  \item Ablations show that structured turn history or scale contributes
    measurable signal beyond naive baselines (chance-level action match,
    empty belief labels, or pilot-scale OOD gaps).
\end{enumerate}

As reported in \Cref{ch:results}, criteria (1) and (4) hold. Criterion (2) does
not hold at current scale. Criterion (3) was not tested and is future work.
Negative results on (2) are treated as scientific findings, not as thesis
failure.

\section{Compute and Reproducibility}

Training runs on CPU or a single GPU. Checkpoints, vocabularies, and evaluation
JSON files are saved under \texttt{checkpoints/} and \texttt{logs/}. The
one-command Phase~2 pipeline is \texttt{scripts/scale\_belief\_research.py}.
The thesis PDF is built from \texttt{thesis/main.tex}.
