\chapter{Results}
\label{ch:results}

\section{Overview}

This chapter reports behavioral cloning and combinatorial generalization results
on Metamon Gen~9 OU human replays. Phase~1 established a turn-only
\texttt{TurnTransformer} baseline on a large encoded subset. Phase~2 exported a
belief-enriched corpus, trained three ablations on the same vocabulary, and
evaluated them under random battle holdout (Split~A) and held-out opponent team
keys (Split~B). All Phase~2 models use an eight-turn context window,
$d=128$, and twenty training epochs unless noted otherwise.

Corpus statistics for the scaled belief export are
69\,411 turns from 2\,000 battles, with 59\,254 supervised move slots and
thousands of item, ability, and terastallization labels on the active opponent.
Gen~9 OU team preview reveals all opponent species at battle start, so species
belief supervision is empty by design. Active-opponent move, item, ability, and
tera type are the primary belief targets.

\section{Phase 1 Baseline (Full Subset)}

The turn-only \texttt{TurnTransformer} was trained on
\texttt{logs/encoded\_metamon\_subset.jsonl} (approximately 255\,000 turns).
On a random 20\% validation split this model reached
\textbf{18.4\%} top-1 action match rate across 988 action classes
(checkpoint \texttt{checkpoints/human\_bc/bc\_best.pt}).
This exceeds the chance rate and confirms that the tokenization pipeline and
hierarchical encoder learn non-trivial structure from human trajectories alone.

\section{Policy Quality (RQ2)}

\subsection{Random Battle Holdout (Split~A)}

\Cref{tab:policy-val} reports validation action accuracy on the belief corpus
after training on 80\% of battles. All three Phase~2 models share the same
encoded file and token vocabulary.

\begin{table}[ht]
  \centering
  \caption{Top-1 action match on random validation battles (Split~A),
    belief corpus, 2\,000 battles, 20 epochs.}
  \label{tab:policy-val}
  \begin{tabular}{lcc}
    \toprule
    Model & Action classes & Val.\ accuracy \\
    \midrule
    TurnTransformer BC (\texttt{human\_bc\_belief}) & 425 & 20.4\% \\
    Matchup BC (\texttt{matchup\_bc}) & 425 & 20.5\% \\
    ActiveBelief (\texttt{active\_belief}) & 425 & 20.2\% \\
    ActiveBelief + fusion & 425 & 19.6\% \\
    \bottomrule
  \end{tabular}
\end{table}

Turn history alone remains the strongest signal for action prediction at this
scale. Explicit team matchup conditioning and auxiliary belief losses did not
improve random-holdout action accuracy relative to the turn-only baseline.
The best checkpoint for \texttt{active\_belief} reached 20.2\% validation
action accuracy while simultaneously training move, item, ability, and tera
heads.

\subsection{Live Evaluation}

Live win-rate experiments on \texttt{poke-env} against \texttt{RandomPlayer} and
frozen BC opponents are planned but not yet included in this draft.
\todo{Insert live battle win rates from \texttt{scripts/run\_bc\_match.py}.}

\section{Belief Inference (RQ1)}

\subsection{Active-Opponent Move Belief}

The \texttt{ActiveBeliefTransformer} predicts four move slots plus item, ability,
and tera type for the currently active opponent. \Cref{tab:belief-move} summarizes
move-slot top-1 accuracy on the validation split.

\begin{table}[ht]
  \centering
  \caption{Move-slot belief accuracy on the validation split (2\,000-battle corpus).}
  \label{tab:belief-move}
  \begin{tabular}{llc}
    \toprule
    Model & History & Val.\ move belief \\
    \midrule
    Linear probe (frozen BC trunk) & 8 turns & 10.5\% \\
    GRU baseline & 8 turns & 17.2\% \\
    ActiveBelief transformer & 8 turns & 28.1\% \\
  \textbf{MLP baseline} & latest turn only & \textbf{35.5\%} \\
    \bottomrule
  \end{tabular}
\end{table}

Scaling the export from 500 to 2\,000 battles raised \texttt{active\_belief} move
belief from 15.5\% to 28.1\%. Classical baselines on the same labels show that a
latest-turn MLP \emph{outperforms} the transformer (35.5\% versus 28.1\%), while a
GRU over eight turns reaches only 17.2\%. Linear probes on the frozen BC trunk
remain at 10.5\% (\Cref{tab:trunk-probe}). Much move information is therefore
readable directly from the current turn tokens; deep causal history provides
diminishing returns for this target at current scale.

\subsection{Species and Set-Combination Beliefs}

Species belief labels are unavailable in Gen~9 OU because team preview exposes
all six opponent species before turn one. Split~C (held-out item, ability, and
move-set tuples) and calibration metrics (ECE, Brier score) are deferred to a
later revision.
\todo{Report item, ability, and tera belief accuracy and Split~C results.}

\section{Combinatorial Generalization (RQ3)}

\subsection{Held-Out Opponent Teams (Split~B)}

Each opponent team is keyed by sorted species names. Training battles cover
1\,615 team keys. Evaluation compares turns from battles whose opponent team
appeared in training (\emph{ID train teams}) against turns from 385 held-out
team keys never seen during training (\emph{OOD teams}). The generalization gap
is ID train team accuracy minus OOD team accuracy.

\Cref{tab:ood-teams} shows results at pilot (500 battles) and scaled
(2\,000 battles) corpus sizes.

\begin{table}[ht]
  \centering
  \caption{Action match under held-out opponent team keys (Split~B).}
  \label{tab:ood-teams}
  \begin{tabular}{lcccc}
    \toprule
    Model & Scale & ID train & OOD teams & Gap \\
    \midrule
    TurnTransformer BC & 500 & 43.3\% & 41.4\% & 1.9\% \\
    Matchup BC & 500 & 31.1\% & 29.3\% & 1.8\% \\
    \addlinespace
    TurnTransformer BC & 2\,000 & 45.2\% & 45.1\% & 0.05\% \\
    Matchup BC & 2\,000 & 42.6\% & 42.1\% & 0.4\% \\
    ActiveBelief & 2\,000 & 44.3\% & 44.4\% & $-$0.03\% \\
    ActiveBelief + fusion & 2\,000 & 43.8\% & 43.3\% & 0.5\% \\
    \bottomrule
  \end{tabular}
\end{table}

On the pilot corpus, all models showed a one to two point gap between ID and OOD
team splits. After scaling to 2\,000 battles, the gap collapsed to near zero for
every architecture. Turn-only BC retained the highest absolute OOD accuracy
(45.1\%). Matchup conditioning did not close the gap further and scored lower in
absolute terms. ActiveBelief matched turn-only OOD performance despite multi-task
belief training.

Random Split~A accuracy ($\approx$20\%) is much lower than Split~B team-conditioned
accuracy ($\approx$42--45\%) because the splits measure different quantities.
Split~A holds out entire battle trajectories. Split~B evaluates action match on
turns grouped by whether the opponent roster was seen during training, which
partially aligns train and test action distributions when species composition is
the primary shift variable.

\section{Architecture Ablations (RQ4)}

Phase~2 implements three incremental ablations on a shared trunk.
\begin{itemize}
  \item \textbf{TurnTransformer BC} uses only turn-history tokens.
  \item \textbf{Matchup BC} adds a \texttt{TeamMatchupEncoder} over both teams'
    six species and fuses the result into the final turn embedding before the
    action head.
  \item \textbf{ActiveBelief} extends Matchup BC with move, item, ability, and
    tera prediction heads trained with masked cross-entropy.
\end{itemize}

At 2\,000 battles, matchup and belief modules did not outperform turn-only BC on
either Split~A or Split~B. This suggests that, at current data and model scale,
the thirteen-token turn representation already encodes most team and belief
information useful for action matching. Explicit structure helps move inference
(RQ1) more than policy metrics (RQ2).

\subsection{Belief Fusion}

We tested concatenating belief softmax outputs (four move slots, item, ability,
tera) to the policy head input (\Cref{ch:model}). After twenty epochs,
\texttt{active\_belief\_fusion} reached \textbf{19.6\%} Split~A validation action
match and \textbf{27.6\%} move belief, versus 20.2\% and 28.1\% for parallel
heads without fusion. OOD team accuracy was 43.3\% (fusion) versus 44.4\%
(without fusion). Feeding explicit belief probabilities into the action head did
not improve imitation at this scale.

Flat $13T$ token transformers and context-window sweeps remain future work.
\todo{Report flat-sequence and context-window ablations.}

\section{Stage B Offline RL}

Conservative offline RL fine-tuning (CQL or AMAGO) initialized from belief-
augmented BC has not been run.
\todo{Insert Stage~B offline RL results if pursued.}

\section{Summary}

\Cref{tab:results-summary} consolidates the scaled Phase~2 numbers. Belief
accuracy refers to move-slot top-1 prediction for \texttt{active\_belief} only.
OOD action match is Split~B OOD team accuracy.

\begin{table}[ht]
  \centering
  \caption{Results summary on the 2\,000-battle belief corpus (20 epochs).}
  \label{tab:results-summary}
  \begin{tabular}{lccc}
    \toprule
    Model & Move belief (val) & OOD action match & Gen.\ gap \\
    \midrule
    TurnTransformer BC & --- & 45.1\% & 0.05\% \\
    Matchup BC & --- & 42.1\% & 0.4\% \\
    ActiveBelief & 28.1\% & 44.4\% & $-$0.03\% \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Relation to Hypotheses}

\textbf{H1 (belief inference).} Mixed. The transformer learns non-trivial move
beliefs and beats the GRU baseline (28.1\% versus 17.2\%), but a latest-turn MLP
reaches 35.5\% on the same split. History-conditioned attention is not required
for strong move prediction when current-turn tokens already expose most clues.

\textbf{H2 (belief usefulness).} Not supported at current scale. Auxiliary
belief loss did not improve action match over turn-only BC on either split.

\textbf{H3 (combinatorial generalization).} Mixed. Scaling data removed the OOD
team gap for all models. Belief and matchup modules did not produce a smaller
gap than the baseline.

\textbf{H4 (offline RL).} Not yet tested.

Reproducibility artifacts are stored under \texttt{checkpoints/} and
\texttt{logs/ood\_comparison.json}. The one-command replication script is
\texttt{scripts/scale\_belief\_research.py}.
